A Higher-Order Approach to Explicit Contexts in LF

### 2015-03-08 paper/paper.tex
Demonstrate the encoding of pi and sigma

\newcommand\naturals{\mathbb{N}}
\newcommand\type{\mathbf{type}}
\newcommand\tyvoid{\mathsf{void}}
\newcommand\tyunit{\mathsf{unit}}
\newcommand\ax{\mathsf{ax}}
\newcommand\ap{\mathsf{ap}}
\newcommand\sortval{\mathsf{val}}
\newcommand\sortexp{\mathsf{exp}}
\newcommand\sortctx{\mathsf{ctx}}
\newcommand\sorttele{\mathsf{tele}}
\newcommand\sortoexp[1]{[#1]\sortexp}
\newcommand\sortj{\mathsf{judg}}
\newcommand\sortoj[1]{[#1]\sortj}
\newcommand\eval[2]{\InputMode{#1}\mathbin{\Rightarrow}\OutputMode{#2}}
\newcommand\thunk[1]{{\color{Gray}\uparrow}#1}
\newcommand\evalrule[1]{\Rightarrow_{#1}}
\newcommand\nobind[1]{{\color{Gray}!}#1}
\newcommand\bind[2]{{\color{Gray}@}[#1]#2}

@@

To motivate this work, let us consider an encoding of Martin-L\"of's
extensional, polymorphic type theory (MLTT~1979) in the logical framework (see
\cite{Martin-Lof-1979} for a full presentation). The syntax contains two sorts,
$\sortval$ (for canonical forms) and $\sortexp$ (for non-canonical forms), and
is defined in the following LF signature: \footnote{LF signatures are presented
  in the style of inference rules for clarity. Object judgement forms have a
  \emph{mode} which expresses which arguments represent inputs, and which
  represent outputs; in our presentation, we color the inputs
  \emph{\InputMode\InputModeColorName}, and the outputs
  \emph{\OutputMode\OutputModeColorName}. Proper assignment of modes may be
used to construe a judgement as having algorithmic content.}

@@

But for terms which involve binding, the rules will become more complicated.
Informally, Martin-L\"of gives the following formation rule for $\mathsf\Pi$
using hypothetico-general judgement:
\[
  \infer[*]{
    \canset{\mathsf{\Pi}(x\in A)B(x)}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }
\]
So we might be tempted to do something similar in the LF, as follows:
\[
  \infer[*']{
    \canset{\typi(A;[x]B(x))}
  }{
    \isset{A} &
    \lfpi{x}\mem{x}{A}\to\isset{B(x)}
  }
\]

@@

>>>>

\newcommand\naturals{\mathbb{N}}
\newcommand\type{\mathbf{type}}
\newcommand\tyvoid{\mathsf{void}}
\newcommand\tyunit{\mathsf{unit}}
\newcommand\ax{\mathsf{ax}}
\newcommand\ap{\mathsf{ap}}
\newcommand\sortval{\mathsf{val}}
\newcommand\sortexp{\mathsf{exp}}
\newcommand\sortctx{\mathsf{ctx}}
\newcommand\sorttele{\mathsf{tele}}
\newcommand\sortoexp[1]{[#1]\sortexp}
\newcommand\sortj{\mathsf{judg}}
\newcommand\sortoj[1]{[#1]\sortj}
\newcommand\eval[2]{\InputMode{#1}\mathbin{\Rightarrow}\OutputMode{#2}}
\newcommand\thunk[1]{{\color{Gray}\uparrow}#1}
\newcommand\evalrule[1]{\Rightarrow_{#1}}
\newcommand\nobind[1]{{\color{Gray}!}#1}
\newcommand\bind[2]{{\color{Gray}@}[#1]#2}
\newcommand\tacitbind[1]{{\color{Gray}@}#1}

@@

To motivate this work, let us consider an encoding of a fragment of
Martin-L\"of's extensional, polymorphic type theory (MLTT~1979) in the logical
framework (see \cite{Martin-Lof-1979} for a full presentation). The syntax
contains two sorts, $\sortval$ (for canonical forms) and $\sortexp$ (for
non-canonical forms), and is defined in the following LF signature:
\footnote{LF signatures are presented in the style of inference rules for
  clarity. Object judgement forms have a \emph{mode} which expresses which
  arguments represent inputs, and which represent outputs; in our presentation,
  we color the inputs \emph{\InputMode\InputModeColorName}, and the outputs
  \emph{\OutputMode\OutputModeColorName}. Proper assignment of modes may be
used to construe a judgement as having algorithmic content.}

@@

But for terms which involve binding, the rules will become more complicated.
Informally, Martin-L\"of gives the following formation rule for $\mathsf\Pi$
using hypothetico-general judgement:
\[
  \infer[*]{
    \canset{(\mathsf{\Pi}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }
\]
So we might be tempted to do something similar in the LF, as follows:
\[
  \infer[*']{
    \canset{\typi(A;[x]B(x))}
  }{
    \isset{A} &
    \lfpi{x}\mem{x}{A}\to\isset{B(x)}
  }
\]

@@

\section{The Definitions of the Types}
It is now possible to give the definitions of the types; recall that we will
only be causing judgements of the following forms to become evident:
$\canset{A}$, $\canmem{M}{A}$, $\eqcanmem{M}{N}{A}$; all other forms of
judgement have been fully explained. The ``spirit'' of Martin-L\"of's type
theory lies in the concept of evaluation of closed terms to canonical form, and
as such, defining a type means only describing how its canonical verifications
behave.

Let us recall Martin-L\"of's informal rules and we will show how they are
encoded in the logical framework. First, the cartesian product of a family of
sets is defined by the following rules in \cite{Martin-Lof-1979}:
\begin{gather*}
  \infer{
    \isset{(\mathsf{\Pi}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }\quad
  \infer{
    \mem{(\lambda x)b}{(\mathsf{\Pi}x\in A)B}
  }{
    \mem{b}{B}\ (\mem{x}{A})
  }\quad
  \infer{
    \eqmem{(\lambda x)b}{(\lambda x)d}{(\mathsf{\Pi}x\in A)B}
  }{
    \eqmem{b}{d}{B}\ (x\in A)
  }
\end{gather*}
%
These rules are adequately encoded in our system as follows:
\begin{gather*}
  \infer{
    \canset{\typi(A;B)}
  }{
    \isset{A} &
    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
  }\quad
  \infer{
    \canmem{\lam(E)}{\typi(A;B)}
  }{
    \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
  }\quad
  \infer{
    \eqcanmem{\lam(E)}{\lam(E')}{\typi(A;B)}
  }{
    \sequent{\cdot,A}{\eqcanmem{\tacitbind{E}}{\tacitbind{E'}}{\tacitbind{B}}}
  }
\end{gather*}

The disjoint union of a family of sets is also easily encoded; first
Martin-L\"of's rules:
\begin{gather*}
  \infer{
    \isset{(\mathsf{\Sigma}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }\quad
  \infer{
    \mem{(a,b)}{(\mathsf{\Sigma}x\in A)B}
  }{
    \mem{a}{A} &
    \mem{b}{[a/x]B}
  }\quad
  \infer{
    \eqmem{(a,b)}{(c,d)}{(\mathsf{\Sigma}x\in A)B}
  }{
    \eqmem{a}{c}{a} &
    \eqmem{b}{d}{[a/x]B}
  }
\end{gather*}
%
and their encoding:
\begin{gather*}
  \infer{
    \canset{\tysg(A;B)}
  }{
    \isset{A} &
    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
  }\quad
  \infer{
    \canmem{\pair(M,N)}{\tysg(A;B)}
  }{
    \mem{M}{A} &
    \mem{N}{B(M)}
  }\quad
  \infer{
    \eqcanmem{\pair(M;N)}{\pair(M';N')}{\typi(A;B)}
  }{
    \eqmem{M}{M'}{A} &
    \eqmem{N}{N'}{B(M)}
  }
\end{gather*}

And what of the elimination rules? The entire point of the verificationist
meaning explanations which pervade \cite{Martin-Lof-1979} is that the
elimination rules will not be part of the definitions of the types, but rather
rules which are admissible, justified by the meanings of the judgements and the
definitions of the types. So, only after having propounded the definitions
above may we entertain \emph{proving} the elimination rules in the logical
framework.

\bigskip
